---
title: "Social Networks (Demography 180)"
editor: source
pagetitle: "Social Networks (Demog 180)"
bibliography: class_networks.bib
#csl: chicago-syllabus.csl
csl: chicago-modified.csl
suppress-bibliography: True
output:
  html_document: 
    css: syllabus.css
    link-external-newwindow: true
  pdf_document: default
---

<!--
NOTES:
- the chicago-syllabus.csl is from Kieran Healy: https://github.com/kjhealy/pandoc-templates/blob/master/csl/chicago-syllabus.csl
- if changing this in the future, you may have to tweak the 'range' values in the part that loads the excel sheet (see grab_syllabus_gsheet in helper.R)

WISHLIST:
- be able to generate pdf version (right now, the table looks wonky - links don't render, NAs, etc)
-->

```{r download-schedule, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

# this file also loads some packages
source('helper.R')

# this file is set so that anyone can read it (but not edit it).
# this means we can avoid annoying authentication requirements
class_url <- "https://docs.google.com/spreadsheets/d/1NbBhyBkNbxTl_MnDGjq5-44N6CKa8ppksMel-3wZ_W4/edit?gid=0#gid=0"

tab <- grab_syllabus_gsheet(class_url)

## deduce all of the states based on the starting date
## THIS ASSUMES THAT
##   - there is a column called 'Date'
##   - the first entry in that column is the date to base the rest of the schedule off of
base_date <- as.Date(tab$Date[1])

next_class <- syllabusr::make_next_class_date(base_date, 'TuTh')
class_date <- syllabusr::make_class_date(base_date, 'TuTh')
```

(Syllabus last updated: `r format(Sys.Date(), "%Y-%B-%d")`)

## Quick links

**Class meetings**: Tuesdays and Thursdays, 2:00-3:30PM, 9 Lewis Hall 
<BR>
**Web**: <https://www.dennisfeehan.org/teaching/demog180-fa2025>  
**Ed page**: <https://edstem.org/us/courses/81974/discussion>  
**Gradescope page**: <https://www.gradescope.com/courses/1084072>  
**Bcourses page**: <https://bcourses.berkeley.edu/courses/1547380>  
**Lecture slides**: <https://drive.google.com/drive/folders/1jgKjkAImWtAzIusbB9mLGhthLyISQHZ2?usp=drive_link>  
<BR>
**Final exam**: TBA   
<BR>

## Staff
   
**Professor Dennis Feehan**, feehan [at] berkeley.edu (but please post questions on Ed)  
Office hours: (see Ed post)  

**TA: Xinghe Pan**  
**TA: Nick Nolte**  

## Overview

The science of social networks focuses on measuring, modeling, and understanding the different ways that people are connected to one another.  In this class, we will use a broad toolkit of theories and methods drawn from the social, natural, and mathematical sciences to learn what a social network is, to understand how to work with social network data, and to illustrate some of the ways that social networks can be useful in theory and in practice. We will see that network ideas are powerful enough to be used everywhere from CDC and UNAIDS, where network models help epidemiologists prevent the spread of HIV, to Silicon Valley, where data scientists use network ideas to build products that enable people all across the globe to connect with one another.


**Please re-check the syllabus frequently; it will be updated as the semester progresses**

```{r show-schedule, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
tabshow <- tab %>%
  mutate(Date = map_chr(row_number(), ~ class_date(.x)),
         # ugly, but working. the contortions are needed to 
         # convert the readable class date into a Date object
         Week = map_dbl(Date, 
                        ~ syllabusr::week_num( mdy(paste(str_replace(.x, ".*,", ""), 
                                                         year(base_date))))))

options(knitr.kable.NA = '')

tabshow %>%
  gt() %>%
  sub_missing(missing_text='') %>%
  fmt_markdown(columns='Theme') %>%
  fmt_markdown(columns='Week') %>%
  fmt_markdown(columns='Demo') %>%
  fmt_markdown(columns='Lab') %>%
  fmt_markdown(columns='Hwk') %>%
  cols_width( Theme ~ pct(10),
              Week ~ pct(5),
              Date ~ pct(5),
              Topic ~ pct(40),
              Demo ~ px(20),
              Lab ~ px(10),
              Hwk ~ px(10))
```



## Detailed modules

<div class="singlespace">

### Introduction to social networks and personal networks {#sec:intro}
	
Required reading: 

* @watts_six_2003 
  * preface-Ch.1
* @easley_networks_2010 
  * Ch.1-Ch.2
* Fischer, [*Still connected*](https://muse.jhu.edu/book/15006), Ch. 2 and 7

### Network structure: foundations {#sec:structure}

Required reading:

* @easley_networks_2010 
  * Ch. 3.1-3.3 (Triadic closure + tie strength)
  * Ch.4.3-4.4 (Affiliation networks)
  * Ch. 3.5 (Social capital)
  * Ch.5.1-5.2 (Positive and negative relationships)
<!-- SHOULD THESE BE PART OF AN ASSIGNMENT? -->
* [Friends you can count on](https://opinionator.blogs.nytimes.com/2012/09/17/friends-you-can-count-on/)
* [Why your friends have more friends than you do](https://www.journals.uchicago.edu/doi/abs/10.1086/229693)

Optional reading:  
  
* @menczer_first_2020 
  * Ch 1.1-1.7 (Introduction to graph theory and `networkx` Python package)
  * Ch 2.2-2.5 (paths and distances; connectedness and components; trees; finding shortest paths)
  * Appendix A: Python tutorial
* the [`networkx` package tutorial](https://networkx.org/documentation/stable/tutorial.html) might be helpful if you want a systematic introduction to `networkx`

### Small worlds and beyond {#sec:smallworlds}

Required readings:

* @watts_six_2003 
  * Ch. 2 (Random networks)
  * Ch. 3 (Small worlds)
  * Ch. 4 (Beyond small worlds; scale-free networks)
  * Ch. 5 (Search in networks)
* @easley_networks_2010 
  * Ch. 20.1-20.2 (Small worlds)
  * Ch. 20.3-20.5 (Search in small worlds)
  * Ch. 18.1-18.5 (Scale-free networks)
* @menczer_first_2020 
  * Sec. 5.4 (Preferential attachment, related to scale-free networks)

Optional readings:

* @menczer_first_2020 
  * Sec. 2.1 ('Birds of a feather' - homophily + assortativity)
  * Sec. 2.6-2.8 (six degrees, clustering coefficient)
  * Sec. 7.4 (Search; especially 7.4.2, Searchability)

### Network structure: advanced {#sec:advancedstructure}

Required reading:

* @menczer_first_2020 
  * Sec. 5.3 (Configuration model; you can skip Exponential Random Graphs)
  * Sec. 6.3 (Community detection; you can skim 6.3.2 and 6.3.3; be sure to read 6.3.4 on Stochastic Block Modeling)

### Simple contagion {#sec:simplecontagion}

Required reading:

* @watts_six_2003 
  * Ch.6
* @easley_networks_2010 
  * Ch. 21.1-21.3 (The SIR epidemic model)

### Concurrency in sexual networks {#sec:concurrency}

Required reading:

  * Sexual networks, concurrency, and HIV
  * @epstein_invisible_2008 
    * Ch.2-4
    
Optional reading    
  
  * @easley_networks_2010 
    * Ch. 21.6
  * NOTE: if you are interested in reading more of the debate over concurrency, [this issue](https://link.springer.com/journal/10461/14/1/page/1) of the journal that Lurie and Rosenthal published in has papers on both sides. (These additional papers are not required reading.)


### Social influence {#sec:socialinfluence}

Required reading:

* @easley_networks_2010 
  * Ch. 16.1-16.2; 
  * parts of 16.3-16.6; 16.7
* @watts_six_2003 
  * Ch. 7 

### Complex contagion {#sec:complexcontagion}

Reading:

* @watts_six_2003 
  * Ch. 8
* @easley_networks_2010 
  * Ch. 19.1-19.6
* [Study says obesity can be contagious](https://www.nytimes.com/2007/07/25/health/25cnd-fat.html)

### Cooperation {#sec:cooperation}

Required reading:

* @easley_networks_2010 
  * Ch. 6.1-6.2 
* @axelrod_evolution_1984 Ch. 1 (pdf in Ed post)



<!--
### Week 15: Guest lectures {#sec:guests}

Reading:

* TBA

-->

::: {.callout-note appearance='simple' icon=false collapse='true'}

## Requirements


### Grades

Your grade will be calculated as a weighted average of several different
class requirements, which are explained in greater detail below.

|  Component | % of grade |
|  ------ | ------ |
|  Labs  (~8; you can drop your lowest score) | 15 |
|  Mini-projects (~4 of these) | 25 |
|  Problem sets (3 of these)   | 10 |
|  Mid-term exam  | 15 |
|  Final exam | 30 |
|  Participation + Quizzes | 5 |


### Lectures

Lectures will introduce and develop key theoretical and technical concepts in the study of social networks.  To illustrate these ideas, some of the lectures will have a live lab / demo component, where  we will interactively discuss and work through an analysis in a Jupyter notebook. These live labs will help us explore and develop intuition about key concepts in the course.

*You are responsible for all of the material covered in lectures, as well as any announcements made there.*

### Required readings

The course readings will include selections from the textbook 
[Networks, Crowds, and Markets](https://www.cs.cornell.edu/home/kleinber/networks-book/) by Easley and Kleinberg: 

* @easley_networks_2010

The Easley and Kleinberg book is freely available online from the author's website. You can also buy a paper copy if you like.

For a couple of topics, I will also assign short readings from the textbook
[A First Course in Network Science](https://www-cambridge-org.libproxy.berkeley.edu/highereducation/books/a-first-course-in-network-science/EE22722F27519D8BB1443C7225C57BAF#contents) by Menczer, Fortunato, and Davis:

@menczer_first_2020

This book is available online through UC Berkeley's library. Menczer et al also has Python code in it, so if you want to see more Python examples, it would be a good place to look.
  
We will also read chapters from popular science books written by leading network researchers, including selections from

* @watts_six_2003
* @epstein_invisible_2008
  
I will post scans of selections of the Epstein book as .pdfs. You should plan to buy
the Watts book, which should not be very expensive.   
  
There will also be selected other readings where they are relevant to our discussions.

The readings serve two purposes: (1) they provide an introduction and reference for key concepts that we will need to study social networks; (2) they illustrate how social network ideas get used in real world research and applications across many different disciplines. You are expected to do the reading before each class. Whenever possible,  PDFs of the readings will be posted on the bCourses site.


### Homeworks and labs

There will be a total of 6 to 8 homeworks, a similar number of labs, and one mini-project.  These assignments are a critical part of the learning you will do in this class: they give you an opportunity to explore the topics we cover in the readings and in lecture on your own. They also give you a chance to practice your writing and your data analysis and programming skills.  Most homeworks and labs will ask you to provide some written arguments and to solve some problems by writing Python code in a Jupyter notebook.  

Labs are graded based on effort; therefore, you can get full credit on a lab even if you do not get all of the answers right. Labs must be handed in on time for full credit. 

Homeworks and are graded on correctness and must be handed in on time for full credit. However, we will drop the homework with the lowest score; thus, you can miss handing in one homework over the course of the semester without it affecting your grade. (Note that you cannot drop the grade for the mini-project.)

The mini-project is like an extended homework that comes after all of the notebook-based homeworks. The goal is to give you a chance to start from scratch with a new network dataset and to demonstrate that you can perform an analysis on the network with minimal hand-holding. It will count as two homeworks.

**Collaboration**: It can be helpful and educational to discuss the assignments with other students in the class, but (1) all of the work should be your own (i.e., you are not allowed to just copy code, answers, or arguments); (2) you should make a note of the names of the other students you worked with when handing your assignments in. Please treat AI tools like ChatGPT like another student: follow rules (1) and (2); that is, don't copy code or text directly from an AI tool and please make a note of any tool you consulted at the top of your assignment.


### Exams

There will be two in-class closed book examinations. The mid-term examination will be held during normal class time in our normal classroom; the timing of this midterm will be designed to assess your mastery of the core concepts in social networks. The final will be held during the final exam period (see the date/time above). The final exam will be cumulative.

### Participation and quizzes

I will sometimes use iClicker to ask interactive questions during lectures. The correctness of responses to these questions will not matter - but they will be used as evidence of attendance.  In some lectures, you may also be asked to participate in discussions and interactive demonstrations. Finally, there will also be a small number (about 2) quizzes on bCourses over the semester. These quizzes will consist of a few multiple choice questions; the goal of these quizzes will be to ensure that you are staying up to date with the reading and lecture materials covered in the class (including guest lectures). I will announce the timing of these quizzes in advance.

:::

::: {.callout-note appearance='simple' icon=false collapse='true'}


## Other class policies  
  
**Religious Accommodations**  
  
Requests to accommodate a student's religious creed by scheduling tests or
examinations at alternative times should be submitted directly to the
instructor. Reasonable common sense, judgment and the pursuit of mutual
goodwill should result in the positive resolution of scheduling conflicts. The
regular campus appeals process applies if a mutually satisfactory arrangement
cannot be achieved.

**Statement on Academic Freedom**  
  
Both students and instructors have rights to academic freedom. Please respect
the rights of others to express their points of view in the classroom.
  
**DSP Accommodations**  
  
The purpose of academic accommodations is to ensure that all students have a fair chance at academic success. Disability, or hardships such as basic needs insecurity, uncertain documentation and immigration status, medical and mental health concerns, pregnancy and parenting, significant familial distress, and experiencing sexual violence or harassment, can affect a student’s ability to satisfy particular course requirements. Students have the right to reasonable academic accommodations, without having to disclose personal information to instructors. For more information about accommodations, scheduling conflicts related to religious creed or extracurricular activities, please see the [Academic Accommodations hub website](https://evcp.berkeley.edu/programs-resources/academic-accommodations-hub#accommodations).
  
**Student Resources**  
  
The Student Learning Center provides a wide range of resources to promote learning and academic success for students. For information regarding these services, please consult the Student Learning Center Website:
https://slc.berkeley.edu/

**Classroom Climate**  
  
We are all responsible for creating a learning environment that is welcoming, inclusive, equitable, and respectful. If you feel that these expectations are not being met, you can consult your instructor(s) or seek assistance from campus resources (see the [Academic Accommodations website](https://evcp.berkeley.edu/programs-resources/academic-accommodations-hub#accommodations)).  
  
**AI**  
  
Your work is expected to reflect your own ideas and should be prepared mostly without reliance on generative AI. Appropriate uses include translation, limited editing, early brainstorming, and formatting. Copying and pasting from an LLM into an assignment is not allowed, even if the results are then edited. If you are unsure if something is appropriate, ask. A good rule of thumb is to consider whether you would feel confident disclosing your process to your GSI or to me. Using AI extensively will be considered plagiarism and reported to Student Conduct.
  
**Academic Integrity**  
  
The high academic standard at the University of California, Berkeley, is reflected in each degree that is awarded. As a result, every student is expected to maintain this high standard by ensuring that all academic work reflects unique ideas or properly attributes the ideas to the original sources.

These are some basic expectations of students with regards to academic integrity:

* Any work submitted should be your own individual thoughts, and should not have been submitted for credit in another course unless you have prior written permission to re-use it in this course from this instructor.
* All assignments must use "proper attribution," meaning that you have identified the original source and extent or words or ideas that you reproduce or use in your assignment. This includes drafts and homework assignments!
* In general, you should not turn in work that was done by an AI tool, such as an LLM (like ChatGPT). If you have any questions, please ask an instructor.
* If you are unclear about expectations, ask your instructor or GSI.
* Do not collaborate or work with other students on assignments or projects unless you have been given permission or instruction to do so.

:::
